---
title: "Attendance and Graduation Rates"
author: "Edosa Odia"
editor: visual
code-fold: true
---

# Introduction

Hello, thank you for taking the time to delve into this independent study, which explores the question of what factors are impacting overall academic performance in New York City’s public high schools. My specific task was to analyze attendance and graduation rates across New York City's school districts from 2018 to 2023. As a New Yorker and an alumnus of the NYC school system, this research is particularly close to my heart.

In this report, we will analyze data on attendance and graduation rates from the nyc doe infohub to uncover trends and draw insights that may inform future educational strategies within the city. Hope you enjoy!

## Relevant Libraries

```{r}
library(tidyverse)
library(ggplot2)
library(scales)
library(corrplot)
library(readxl)
library(DT)
library(sf)
library(gt)
```

These are all the packages necessary for this analysis. If you do not have any of these libraries installed use `install.packages()` to install what you may need.

## Loading the data

```{r}
get_github_excel_all_sheets <- function(file_path) {
  
  # Construct the full GitHub URL to the raw version of the file
  base_url <- "https://raw.githubusercontent.com/"
  full_url <- paste0(base_url, file_path)
  
  # Temporarily download the file
  temp_file <- tempfile(fileext = ".xlsx")
  download.file(full_url, temp_file, mode = "wb")  
  sheet_names <- excel_sheets(temp_file)
  
  sheets_data <- lapply(sheet_names, function(sheet) {
    read_excel(temp_file, sheet = sheet)
  })
  
  # Return the list of data frames
  return(sheets_data)
}

#Github paths

file_path <- "Edodia/STA9750-2024-FALL/main/public-district-attendance-results-2018-2023.xlsx"
file_path2 <- "Edodia/STA9750-2024-FALL/main/2023-graduation-rates-public-district.xlsx"
all_sheets_data <- suppressWarnings(get_github_excel_all_sheets(file_path))
all_sheets_data2 <- suppressWarnings(get_github_excel_all_sheets(file_path2))

Attendance_ALL <- all_sheets_data[[2]]
Attendance_SWD <- all_sheets_data[[3]]
Attendance_Ethnicity <- all_sheets_data[[4]]
Attendance_Gender <- all_sheets_data[[5]]
Attendance_Poverty <- all_sheets_data[[6]]
Attendance_ELL <- all_sheets_data[[7]]
Attendance_STH <- all_sheets_data[[8]]
Graduation_Rates_All <- all_sheets_data2[[2]]
Graduation_Rates_SWD <- all_sheets_data2[[5]]
Graduation_Rates_ELL <- all_sheets_data2[[3]]
Graduation_Rates_Ethnicity <- all_sheets_data2[[6]]
Graduation_Rates_Gender <- all_sheets_data2[[7]]
Graduation_Rates_Poverty <- all_sheets_data2[[8]]
```

This R function retrieves Excel files from GitHub and loads all sheets directly into your R environment. It constructs a URL for the raw file, downloads it temporarily, and then reads each sheet into a separate data frame using the `readxl`package. Ensure `readxl` is installed and loaded before running this function. The datasets include information on studants with disablities (SWD), ethnicity, gender, english language learner status (ELL). Only the attendance dataset includes information on students in temporary housing (STH).

## Initial Cleaning of Attendance Data

```{r}
#Specifiying data is within grades 12 for attendance to match graduation data
filter_high_school_grades <- function(data) {
  filtered_data <- data[data$`Grade` %in% c("12"), ]
  return(filtered_data)
}
Attendance_Poverty <- filter_high_school_grades(Attendance_Poverty)
Attendance_Ethnicity <- filter_high_school_grades(Attendance_Ethnicity)
Attendance_Gender <- filter_high_school_grades(Attendance_Gender)
Attendance_ELL <- filter_high_school_grades(Attendance_ELL)
Attendance_ALL <- filter_high_school_grades(Attendance_ALL)
Attendance_STH <- filter_high_school_grades(Attendance_STH)
Attendance_SWD <- filter_high_school_grades(Attendance_SWD)

```

In an effort to enhance the relevance and precision of our analysis, I have refined the scope of this section of our project to concentrate exclusively on twelfth-grade students. This decision aligns the attendance data directly with our graduation dataset, enabling a focused examination of student outcomes at the end of their high school journey.

## 

## Attendance Analysis 

```{r}
#Attendance Analysis
# Identifying the top five worst and best districts in NYC for attendance
attendance_summary <- Attendance_ALL %>%
  group_by(District) %>%
  summarise(Average_Attendance = mean(`% Attendance`, na.rm = TRUE)) %>%
  arrange(desc(Average_Attendance))


print(attendance_summary)
top_districts <- attendance_summary %>%
  top_n(5, wt = Average_Attendance) %>%
  arrange(desc(Average_Attendance))

# Plot 
ggplot(top_districts, aes(x = reorder(District, Average_Attendance), y = Average_Attendance, fill = as.factor(District))) +
  geom_col(show.legend = FALSE) +  
  geom_text(aes(label = sprintf("%.2f%%", Average_Attendance)), 
            position = position_stack(vjust = 1.05),  
            color = "black", size = 3.5) +  
  scale_fill_brewer(palette = "Paired") +  
  labs(title = "Top 5 Districts by Average Attendance", x = "District", y = "Average Attendance (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Now, lets get to the meat of this project. As you see here this a quick overview of the five districts with the highest overall average attendance. In order it is district: 26 (89.41%), 24 (88.93%), 4 (88.85%), 28 (88.17%), 22 (88.15%). This is great! However, it is still a bit lower than I expected. I believed that the top five would have averages in the 90% range.

```{r}
#What are the bottom 5
bottom_districts <- attendance_summary %>%
  arrange(Average_Attendance) %>%
  head(5)  

# Print to check
print(bottom_districts)
ggplot(bottom_districts, aes(x = reorder(District, Average_Attendance), y = Average_Attendance, fill = as.factor(District))) +
  geom_col(show.legend = FALSE) +  
  geom_text(aes(label = sprintf("%.2f%%", Average_Attendance)), 
            position = position_stack(vjust = 1.05),  
            color = "black", size = 3.5) +  
  scale_fill_brewer(palette = "Paired") + 
  labs(title = "Bottom 5 Districts by Average Attendance", x = "District", y = "Average Attendance (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  
```

Inversely this visualization shows the bottom five districts and this is absolutely jarring. Not one of these five districts are above an 80% attendance rate, if these students were graded solely on their ability to attend school they would maybe get a C+. The list of the bottom five districts by average attendance from highest to lowest is: district 32 (79.03%), 18 (78.65%), 15 (78.63%), 23 (77.34%), 16 (69.19%).

## 

## Attendance District Map

```{r}
#Now ploting attendance rates using the nyc shapefiles
#Initial setup
if(!file.exists("nysd_24d.zip")){
  download.file("http://s-media.nyc.gov/agencies/dcp/assets/files/zip/data-tools/bytes/nysd_24d.zip", 
                destfile="nysd_24d.zip", method = "curl")
}


td <- tempdir(); 
zip_contents <- unzip("nysd_24d.zip", 
                      exdir = td)

fname_shp <- zip_contents[grepl("shp$", zip_contents)]
nyc_sf <- read_sf(fname_shp)

ggplot(nyc_sf, 
       aes(geometry=geometry, 
           fill = Shape_Area)) + 
  geom_sf()

colnames(nyc_sf)
nyc_sf <- nyc_sf %>%
  dplyr::rename(District = SchoolDist)
nyc_sf$District <- as.character(nyc_sf$District)
attendance_summary$District <- as.character(attendance_summary$District)
nyc_sf_attendance <- merge(nyc_sf, attendance_summary, by = "District", all.x = TRUE)
#now plot
ggplot(nyc_sf_attendance, aes(geometry = geometry, fill = Average_Attendance)) +
  geom_sf() +  
  geom_sf_text(aes(label = District), size = 3, colour = "white", check_overlap = TRUE) +  
  scale_fill_gradient(low = "red", high = "blue",  
                      name = "Average Attendance from 2018 to 2023", 
                      labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Map of NYC School Districts by Average Attendance", 
       subtitle = "District labels are shown on the map") +
  theme_minimal() +
  theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5))
```

This is map that shows all of the districts across the city from 2018 to 2023. The lower the average attendance is the redder the district is and the higher the average attendance is the bluer it is. It seems that the worst attendance rates is concentrated in Bedstuy, the Bronx and the Lower East Side. While, Queens, Staten Island, and the Upper East Side generally show a higher average attendance rate.

## Analysis of Attendance by Demographic Factors

### ELL Status

```{r}
#ELL 
# Separate attendance data for ELL and Not ELL for the test
attendance_ell <- Attendance_ELL$`% Attendance`[Attendance_ELL$Category == "ELL"]
attendance_not_ell <- Attendance_ELL$`% Attendance`[Attendance_ELL$Category == "Not ELL"]

# Perform a t-test
t_test_result <- t.test(attendance_ell, attendance_not_ell)
print(t_test_result)
#The Welch t-test results strongly suggest that non-ELL students attend school at a higher rate than ELL students.
ggplot(Attendance_ELL, aes(x = Category, y = `% Attendance`, fill = Category)) +
  geom_boxplot() +
  labs(title = "Attendance Rates by ELL Status",
       x = "Category",
       y = "Attendance Rate (%)") +
  scale_fill_brewer(palette = "Set1") +  
  theme_minimal() +
  theme(legend.position = "none") 

```

I wanted to do something a little bit different and conduct a Welch t-test in order to analyze the difference between means of the ELL and not ELL students. The Welch t-test results, suggests that non-ELL students(84%) attend school at a higher rate than ELL students (76%), pointing to a potential disparity in attendance that might be influenced by language barriers.

### Gender 

```{r}
Attendance_Gender <- na.omit(Attendance_Gender)
Attendance_Gender$`% Attendance` <- gsub("s", "", Attendance_Gender$`% Attendance`)  # Remove 's'
Attendance_Gender$`% Attendance` <- as.numeric(Attendance_Gender$`% Attendance`)  
str(Attendance_Gender$`% Attendance`)
Attendance_Gender <- na.omit(Attendance_Gender)
summary(Attendance_Gender$`% Attendance`)
# Mean averages
average_attendance_gender <- Attendance_Gender %>%
  group_by(Category) %>%
  summarise(Average_Attendance = mean(`% Attendance`, na.rm = TRUE))

print(average_attendance_gender)
ggplot(average_attendance_gender, aes(x = Category, y = Average_Attendance, fill = Category)) +
  geom_bar(stat = "identity", width = 0.7) +  
  geom_text(aes(label = sprintf("%.2f%%", Average_Attendance)), 
            position = position_stack(vjust = 1.05),  
            color = "black", size = 3.5) +  
  scale_fill_brewer(palette = "Pastel1") +  
  labs(title = "Average Attendance Rates by Gender",
       x = "Gender Category",
       y = "Average Attendance Rate (%)") +
  theme_minimal() +  
  theme(legend.position = "none",  
        axis.text.x = element_text(angle = 45, hjust = 1))

```

This bar chart shows that males and females attend school at a similar rate with women beating men by a small 1% margin. Unfortunately, there is not enough data on non-binary people so they have been left out of this study.

### Students with Disabilities

```{r}
#SWD attendance
Attendance_SWD <- na.omit(Attendance_SWD)
average_attendance_swd <- Attendance_SWD %>%
  group_by(Category) %>%
  summarise(Average_Attendance = mean(`% Attendance`, na.rm = TRUE))
print(average_attendance_swd)
swd_dt <- datatable(average_attendance_swd, options = list(pageLength = 5, autoWidth = TRUE), 
                    filter = 'top',  
                    rownames = FALSE)  
swd_dt
```

Students with disabilities attend school at a lower rate compared to their non disabled counterparts by a margin of 5%. This may be due to any number of factors such as lack of aid to students with disabilities or even possible bullying.

### Students in Temporary Housing

```{r}
#STH
Attendance_STH <-na.omit(Attendance_STH)
average_attendance_sth <- Attendance_STH %>%
  group_by(Category) %>%
  summarise(Average_Attendance = mean(`% Attendance`, na.rm = TRUE))
print(average_attendance_sth)
sth_dt <- datatable(average_attendance_sth, options = list(pageLength = 5, autoWidth = TRUE), 
                    filter = 'top',  
                    rownames = FALSE) 
sth_dt
```

Students in temporary housing also do not attend school as much as students with permanent housing by a margin of 6%. This could have many different explanations as to why this may be such as lack of transportation, lack of resources, social and emotional issues, and general instability.

### Poverty

```{r}
#Poverty
Attendance_Poverty <-na.omit(Attendance_Poverty)
average_attendance_pov <- Attendance_Poverty %>%
  group_by(Category) %>%
  summarise(Average_Attendance = mean(`% Attendance`, na.rm = TRUE))
print(average_attendance_sth)
pov_dt <- datatable(average_attendance_pov, options = list(pageLength = 5, autoWidth = TRUE), 
                    filter = 'top',  
                    rownames = FALSE)
pov_dt
```

Interestingly though both students who are in poverty and those who are not are above an 80% average attendance rate with the margin being only 4%. However, students in poverty attend school less and the explanations for this is similar to the explanation for students in temporary housing.

### Ethnicity 

```{r}
#Ethnicity
Attendance_Ethnicity <- na.omit(Attendance_Ethnicity)
Attendance_Ethnicity$`% Attendance` <- gsub("s", "", Attendance_Ethnicity$`% Attendance`)  
Attendance_Ethnicity$`% Attendance` <- as.numeric(Attendance_Ethnicity$`% Attendance`)  
average_attendance_ethnicity <- Attendance_Ethnicity  %>%
  group_by(Category) %>%
  summarise(Average_Attendance = mean(`% Attendance`, na.rm = TRUE))

attendance_eth_plot <- ggplot(average_attendance_ethnicity, aes(x = Category, y = Average_Attendance, fill = Category)) +
  geom_bar(stat = "identity", width = 0.7) +  
  geom_text(aes(label = sprintf("%.2f%%", Average_Attendance)), position = position_stack(vjust = 1.05),
            color = "black", size = 3.5) + 
  scale_fill_brewer(palette = "Paired") + 
  labs(title = "Average Attendance Rates by Ethnicity",
       x = "Ethnic Category",
       y = "Average Attendance Rate (%)") +
  theme_minimal() +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

attendance_eth_plot

```

As you can see here Asian students attend school at the highest rate out of all groups while Black and Hispanic students attend school the least. The Other group is a mixture of native american students and mixed race students. There are many possible explanations of this such as cultural differences or resource differences within each demographic group,

## Graduation!! Someone call Kanye West!

```{r}
#Some data cleaning and modification
filter_by_cohort_year <- function(data) {
  # Define the cohort years to include
  valid_cohorts <- c("2014", "2015", "2016", "2017", "2018")
  
  # Filter the data to include only the specified cohorts
  filtered_data <- data[data$`Cohort Year` %in% valid_cohorts, ]
  
  return(filtered_data)
}
Graduation_Rates_All <- filter_by_cohort_year(Graduation_Rates_All)
Graduation_Rates_ELL <- filter_by_cohort_year(Graduation_Rates_ELL)
Graduation_Rates_Poverty <- filter_by_cohort_year(Graduation_Rates_Poverty)
Graduation_Rates_Gender <- filter_by_cohort_year(Graduation_Rates_Gender)
Graduation_Rates_SWD <- filter_by_cohort_year(Graduation_Rates_SWD)

```

This section focuses on cleaning and filtering the data. I selected these specific cohorts as we will be able to correctly match it to the attendance data. Note that the cohort year references the year a student entered high school, so a student of the 2014 cohort would theoretically graduate in 2018.

## Graduation District Map

```{r}
#Now graduation top 5 and bottom 5 districts
graduation_summary <- Graduation_Rates_All %>%
  group_by(District) %>%
  summarise(Average_Graduation = mean(`% Grads`, na.rm = TRUE)) %>%
  arrange(desc(Average_Graduation))

# View the summary
print(graduation_summary)
top_districts <- graduation_summary  %>%
  top_n(5, wt = Average_Graduation) %>%
  arrange(desc(Average_Graduation))
nyc_sf_grads <- merge(nyc_sf, graduation_summary, by = "District", all.x = TRUE)
nyc_sf_grads$District <- as.character(nyc_sf_grads$District)

# Create the map with labels
graduation_map <- ggplot(nyc_sf_grads, aes(geometry = geometry, fill = Average_Graduation)) +
  geom_sf() +  # Draw the shapes
  geom_sf_text(aes(label = District), size = 3, colour = "white", check_overlap = TRUE, position = position_nudge(y = 0.02)) +  # Add district labels
  scale_fill_gradient(low = "red", high = "blue",  
                      name = "Average Graduation from 2018 through 2024",
                      labels = percent_format(accuracy = 1)) +
  labs(title = "Map of NYC School Districts by Average Graduation Rate",
       subtitle = "Graduation rates from 2018 through 2024") +
  theme_minimal() +
  theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5))
graduation_map

```

I hope you are enjoying all of these colorful maps! This is map showing all of the districts from 2018 to 2024. The redder it is the lower the graduation rates and the bluer it is the higher the graduation rate. It seems that the Bronx and Bedstuy are having a problem when it comes to graduation rates while Queens is excelling.

## Top 5 Graduation Rates

```{r}
#top and bottom 5 grad rates per district
top_5_grad_districts <- head(graduation_summary,5)
bottom_5_grad_districts <- tail(graduation_summary,5)

top_5_grad_plot<- ggplot(top_5_grad_districts, aes(x = reorder(as.factor(District), Average_Graduation), y = Average_Graduation, fill = as.factor(District))) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = sprintf("%.1f%%", Average_Graduation)), vjust = -0.5, color = "black") +  
  scale_fill_brewer(palette = "Blues") +
  labs(title = "Top 5 Districts by Average Graduation Rate",
       x = "District",
       y = "Average Graduation Rate (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
top_5_grad_plot
```

This bar chart shows that the districts with the highest graduation rates. Go district 26! They had the highest graduation rate and attendance rate out of all districts.

## Bottom 5 Graduation Rates

```{r}
bottom_5_grad_plot <- ggplot(bottom_5_grad_districts, aes(x = reorder(as.factor(District), -Average_Graduation), y = Average_Graduation, fill = as.factor(District))) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = sprintf("%.1f%%", Average_Graduation)), vjust = -0.5, color = "black") +  
  scale_fill_brewer(palette = "Reds") +
  labs(title = "Bottom 5 Districts by Average Graduation Rate",
       x = "District",
       y = "Average Graduation Rate (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
bottom_5_grad_plot
```

This is terrible. The average graduation rates range from 71.4% to 56.3% for the bottom five districts. District 16, 23, and 18 were also all in the bottom five in attendance rate. Is there a correlation?

## Analysis of Attendance by Demographic Factors

### Ethnicity

```{r}
Graduation_Rates_Ethnicity <- na.omit(Graduation_Rates_Ethnicity)
Graduation_Rates_Ethnicity$`% Grads` <- gsub("s", "", Graduation_Rates_Ethnicity$`% Grads`)  
Graduation_Rates_Ethnicity$`% Grads` <- as.numeric(Graduation_Rates_Ethnicity$`% Grads`) 
average_grad_ethnicity <- Graduation_Rates_Ethnicity  %>%
  group_by(Category) %>%
  summarise(Average_Graduation = mean(`% Grads`, na.rm = TRUE))

grad_ethn_plot <- ggplot(average_grad_ethnicity, aes(x = Category, y = Average_Graduation, fill = Category)) +
  geom_bar(stat = "identity", width = 0.7) +  
  geom_text(aes(label = sprintf("%.2f%%", Average_Graduation)), position = position_stack(vjust = 1.05),
            color = "black", size = 3.5) +  
  scale_fill_brewer(palette = "Paired") +  
  labs(title = "Average Graduation Rates by Ethnicity",
       x = "Ethnic Category",
       y = "Average Graduation Rate (%)") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
grad_ethn_plot
```

This chart is very similar to the ethnicity chart for attendance. Asians and multi-racial students have the highest graduation rates while black, Hispanic, and Native Americans have the lowest graduation rates. Luckily, this dataset had more information on mixed race and Native American kids.

### Poverty

```{r}
Graduation_Rates_Poverty <- na.omit(Graduation_Rates_Poverty)
average_grad_pov <- Graduation_Rates_Poverty  %>%
  group_by(Category) %>%
  summarise(Average_Graduation = mean(`% Grads`, na.rm = TRUE))
average_grad_pov
grad_pov_dt <- datatable(average_grad_pov, options = list(pageLength = 5, autoWidth = TRUE), 
                            filter = 'top',  
                            rownames = FALSE) 
grad_pov_dt
```

Interestingly though economically disadvantaged and not economically disadvantaged students graduate at a similar rate with a margin of less than 1%.

### English Language Learner Status

```{r}
Graduation_Rates_ELL <-na.omit(Graduation_Rates_ELL)
average_grad_ell <- Graduation_Rates_ELL  %>%
  group_by(Category) %>%
  summarise(Average_Graduation = mean(`% Grads`, na.rm = TRUE))
average_grad_pov
grad_ell_dt <- datatable(average_grad_ell, options = list(pageLength = 5, autoWidth = TRUE), 
                         filter = 'top',  
                         rownames = FALSE) 
grad_ell_dt
```

ELL students have a very low graduation rate of 52% and those who were never ELL have a graduation rate of 81%. However, what is truly interesting is that students who managed to exit out of the ELL program fared the best out of all three categories.

### Students With Disabilities

```{r}
Graduation_Rates_SWD <-na.omit(Graduation_Rates_SWD)
average_grad_swd <- Graduation_Rates_SWD  %>%
  group_by(Category) %>%
  summarise(Average_Graduation = mean(`% Grads`, na.rm = TRUE))
average_grad_swd
grad_swd_dt <- datatable(average_grad_swd, options = list(pageLength = 5, autoWidth = TRUE), 
                         filter = 'top',  
                         rownames = FALSE)
grad_swd_dt
```

This is also a scary statistic. Students who do not have a disability fare much better than students who do by a margin of 27%.

### Gender

```{r}
Graduation_Rates_Gender <-na.omit(Graduation_Rates_Gender)
Graduation_Rates_Gender$`% Grads` <- gsub("s", "", Graduation_Rates_Gender$`% Grads`)  
Graduation_Rates_Gender$`% Grads` <- as.numeric(Graduation_Rates_Gender$`% Grads`) 
Graduation_Rates_Gender <- Graduation_Rates_Gender[!is.nan(Graduation_Rates_Gender$`% Grads`), ]

average_grad_gender <- Graduation_Rates_Gender  %>%
  group_by(Category) %>%
  summarise(Average_Graduation = mean(`% Grads`, na.rm = TRUE))
average_grad_gender
grad_gen_dt <- datatable(average_grad_gender, options = list(pageLength = 5, autoWidth = TRUE), 
                         filter = 'top',  
                         rownames = FALSE)
grad_gen_dt
```

Women beat out men in both attendance and graduation rates. This time by a huge margin of almost 9%. Is it true girls are smarter than boys?

## Is Attendance Correlated to Graduation Rates?

```{r}
#Answering the true question!
combined_data <- merge(Graduation_Rates_All, Attendance_ALL, by = "District")

# Calculate correlation between `% Grads` and `% Attendance`
correlation_result <- cor(combined_data$`% Grads`, combined_data$`% Attendance`, use = "complete.obs")
print(correlation_result)
#correlation of .608 thats really strong
# Visualizing the correlation with a scatter plot
final_corr <- ggplot(combined_data, aes(x = `% Attendance`, y = `% Grads`)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Correlation between Attendance and Graduation Rates for Cohorts 2014-2018",
       x = "Attendance Rate (%)",
       y = "Graduation Rate (%)") +
  theme_minimal()
final_corr

```

My analysis confirms a significant positive correlation between student attendance rates and graduation rates across New York City schools, with a correlation coefficient of 0.608. This indicates a clear relationship: as attendance increases, graduation rates tend to increase as well.

# Conclusion

This study has thoroughly examined the dynamics between student attendance and graduation rates across various demographic groups within New York City’s educational landscape from 2018 to 2023. My findings indicate that attendance is a fundamental driver of student success and graduation likelihood.

### Key Takeaways & Recommendations:

-   Poverty and Students in Temporary Housing: Students in temporary housing attend school at a rate of 77% which is lower than their peers. This can be remedied by stronger public assistance programs and free mental health and wellness checkups.

-   ELL Students: ELL students also need help. They graduate at a rate close to 50% however, those that do manage to leave the program graduate at a rate of 90%. I would recommend more funding to the ELL program as it is clear that language barriers is detrimental to the success of students.

-   Districts 16, 23, and 18 which are BedStuy, Brownsville, and Bushwick need serious attention. They rank at the lowest in graduation rates and attendance rates. This may be contributed to the poverty in the areas. Regardless though it is obvious that academic outreach programs is necessary for these neighborhoods.

    Thank you I hoped you enjoyed.
